\chapter{Introduction}

In this thesis we present a detailed study of two applications of non-Abelian gauge theories to Beyond the Standard Model physics.
The Standard Model (SM) of particle physics is very successful in describing experimental data. Nevertheless, we know that it is just an effective theory, valid only up to an ultraviolet cutoff. 
The question of what kind of new physics may constitute extensions of the Standard Model is very fascinating, and a large part of the high-energy-physics community is dedicated to uncover what lies beyond it. 
%In this thesis, we consider two deeply related topics belonging to Beyond the Standard Model research: composite Higgs models and the conformal window.

%In the first study, we address the question of mass generation for SM particles. 
Composite Higgs models represent an important branch of Beyond the Standard Model physics. They address the question of mass generation for the SM particles,
aiming to provide a more theoretically consistent framework than the SM Higgs mechanism. 
Almost all phenomenologically viable composite Higgs models are based on gauge theories with  ``walking'' dynamics. These theories have an approximate infrared fixed point, and are believed to exist in theory space close to the lower boundary of the so-called conformal window, where models exhibit large-distance conformality, due to an infrared fixed point in the renormalisation group flow. 
Many lattice studies have been devoted to the search of ``walking'' theories in the last decade, which has proven to be a very difficult task.
In this thesis, we consider a novel approach to mass generation, dubbed ``fundamental partial compositeness'' \cite{Sannino:2016sfx} in which (coloured) fundamental scalar fields are introduced.  No ``walking'' dynamics is needed in this case, and the underlying gauge theory features a QCD-like running coupling. Composite Higgs models based on fundamental partial compositeness are able to provide a complete theory of flavour, alternative to the SM Higgs mechanism. This is in contrast to traditional models of partial compositeness, where, due to model building difficulties, the focus is only on the top sector.

The first study presented in this thesis is a lattice analysis of the SU(2) gauge theory with two fundamental fermion flavours and one fundamental coloured scalar field. This model is intended as a minimal scenario for testing the non-perturbative properties of models of  fundamental partial compositeness. 
Lattice results have the important role of providing all the information which cannot be obtained in perturbation theory, among which the flavour symmetry breaking pattern and the spectrum of massive resonances.
In the second study, we explore the thermodynamic properties of theories in the perturbative region of the conformal window. In particular, we study transport coefficients, which describe the near-equilibrium dynamics of the theory.
A very interesting aspect of this analysis is that it provides a framework where the perturbative results for transport coefficients are trustable at both high and low temperatures. In typical applications to strongly coupled gauge theories, the perturbative approach holds exclusively in the high-temperature regime, while in our analysis, due to the perturbative nature of the infrared fixed point, perturbation theory can be applied both in the high-temperature regime and in the infrared.


This thesis is organised as follows: in chapter \ref{LFT}, we give a general introduction to lattice field theory, mostly focused on the tools needed for the lattice study presented in this thesis. In chapter \ref{GFS}, after introducing composite Higgs models and the fundamental partial compositeness mechanism, we present the results of our lattice study of the SU(2) gauge theory with two fundamental fermions and one fundamental scalar. In chapter \ref{CW}, we present our study of the transport coefficients of theories in the conformal window. Before discussing our results, we provide the definitions of the transport coefficients under analysis, and we describe the method used by the authors of \cite{Arnold:2000dr} to obtain perturbative estimates of the transport coefficients of a hot gauge theory, which constitute the basis of our computations.
In appendix, some general definitions and additional material can be found.